Imperative Programming with Variational Effects

### 2018-07-29 paper/paper.tex
edit variational semantics

\begin{mathpar}
\BigN \and
\BigVar \and
\BigAdd \and
\BigB \and
\BigNot \and
\BigAnd \and
\BigLess
\end{mathpar}

@@

\def \BigVNum {\infer [VA-Num] { } {(C,S,n) \Downarrow_{VA} n}}
\def \BigVRef {\infer [VA-Ref] { } {(C,S,x) \Downarrow_{VA} S(C,x)}}
\def \BigVAdd {\infer [VA-Add] {(C,S,e) \Downarrow_{VA} v \\ (C,S,e') \Downarrow_{VA} v'} {(C,S,e+e') \Downarrow_{VA} \mathit{liftV2}(+,v,v')}}
\def \BigVAChcOne {\infer [VA-Chc1] {\mathit{sat}(C \wedge A) \\ \mathit{sat}(C \wedge \neg A) \\ (C \wedge A,S,e) \Downarrow_{VA} v \\ (C \wedge \neg A,S,e') \Downarrow_{VA} v'} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VA} \tagtree{A}{v}{v'}}}
\def \BigVAChcTwo {\infer [VA-Chc2] {\mathit{unsat}(C \wedge A) \\ \mathit{sat}(C \wedge \neg A) \\ (C \wedge \neg A,S,e') \Downarrow_{VA} v'} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VA} \tagtree{A}{\nothing}{v'}}}
\def \BigVAChcThree {\infer [VA-Chc3] {\mathit{sat}(C \wedge A) \\ \mathit{unsat}(C \wedge \neg A) \\ (C \wedge A,S,e) \Downarrow_{VA} v} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VA} \tagtree{A}{v}{\nothing}}}
\def \BigVAChcFour {\infer [VA-Chc4] {\mathit{unsat}(C \wedge A) \\ \mathit{unsat}(C \wedge \neg A)} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VA} \nothing}}

\def \BigVB {\infer [VB-Bool] { } {(C,S,b) \Downarrow_{VB} b}}
\def \BigVNot {\infer [VB-Not] {(C,S,e) \Downarrow_{VB} v} {(C,S,\text{not}\ e) \Downarrow_{VB} \mathit{liftV}(\neg,v)}}
\def \BigVAnd {\infer [VB-And] {(C,S,e) \Downarrow_{VB} v \\ (C,S,e') \Downarrow_{VB} v'} {(C,S,e\ \text{and}\ e') \Downarrow_{VB} \mathit{liftV2}(\wedge,v,v')}}
\def \BigVLess {\infer [VB-Less] {(C,S,e) \Downarrow_{VA} v \\ (C,S,e') \Downarrow_{VA} v'} {(C,S,e<e') \Downarrow_{VB} \mathit{liftV2}(<,v,v')}}
\def \BigVBChcOne {\infer [VB-Chc1] {\mathit{sat}(C \wedge A) \\ \mathit{sat}(C \wedge \neg A) \\ (C \wedge A,S,e) \Downarrow_{VB} v \\ (C \wedge \neg A,S,e') \Downarrow_{VB} v'} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VB} \tagtree{A}{v}{v'}}}
\def \BigVBChcTwo {\infer [VB-Chc2] {\mathit{unsat}(C \wedge A) \\ \mathit{sat}(C \wedge \neg A) \\ (C \wedge \neg A,S,e') \Downarrow_{VB} v'} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VB} \tagtree{A}{\nothing}{v'}}}
\def \BigVBChcThree {\infer [VB-Chc3] {\mathit{sat}(C \wedge A) \\ \mathit{unsat}(C \wedge \neg A) \\ (C \wedge A,S,e) \Downarrow_{VB} v} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VB} \tagtree{A}{v}{\nothing}}}
\def \BigVBChcFour {\infer [VB-Chc4] {\mathit{unsat}(C \wedge A) \\ \mathit{unsat}(C \wedge \neg A)} {(C,S,\tagtree{A}{e}{e'}) \Downarrow_{VB} \nothing}}

\begin{mathpar}
\BigVNum \and
\BigVRef \and
\BigVAdd \and
\BigVAChcOne \and
\BigVAChcTwo \and
\BigVAChcThree \and
\BigVAChcFour \and
\BigVB \and
\BigVNot \and
\BigVAnd \and
\BigVLess \and
\BigVBChcOne \and
\BigVBChcTwo \and
\BigVBChcThree \and
\BigVBChcFour
\end{mathpar}

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVErr {\infer [VS-Err] { } {(C,S,E,\T,s) \Downarrow_{VS} (S,E,\T)}}
\def \BigVSkip {\infer [VS-Skip] { } {(C,S,E,F,\text{skip}) \Downarrow_{VS} (S,E,F)}}
\def \BigVAssn {\infer [VS-Assn] {(\GetCtx{C}{E},S,e) \Downarrow_{VA} v} {(C,S,E,\F,x \coloneqq e) \Downarrow_{VS} (S[(\GetCtx{C}{E},x) \rightarrow v],E,\F)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C,S',E',F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,s\ \text{seq}\ s') \Downarrow_{VS} (S'',E'',F')}}
\def \BigVIfOne {\infer [VS-If1] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'}) \\ (C \wedge \mathit{whenFalse}(v),S',E',\F) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S'',E'',F \wedge F')}}
\def \BigVIfTwo {\infer [VS-If2] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E}) \\ (C \wedge \mathit{whenFalse}(v),S,E,\F) \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfThree {\infer [VS-If3] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S',E',F)}}
\def \BigVIfFour {\infer [VS-If4] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenFalse}(v)}{E'})} {(C,S,E,\F,\text{if}\ e\ \text{then}\ s\ \text{else}\ s') \Downarrow_{VS} (S,E,\F)}}
\def \BigVWhileT {\infer [VS-WhileT] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C \wedge \mathit{whenTrue}(v),S',E',F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S'',E'',F')}}
\def \BigVWhileF {\infer [VS-WhileF] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E})} {(C,S,E,\F,\text{while}\ e\ \text{do}\ s) \Downarrow_{VS} (S,E,\F)}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{h:t},S,e) \Downarrow_{VA} v} {(C,S,h:t,\F,\text{throw}\ e) \Downarrow_{VS} (S,((C,v):h):t,\T)}}
\def \BigVTry {\infer [VS-Try] {(C,S,[]:E,\F,s) \Downarrow_{VS} (S',[]:E,\F)} {(C,S,E,\F,\text{try}\ s\ \text{catch}\ x\ \text{in}\ s') \Downarrow_{VS} (S',E,\F)}}
\def \BigVCatch {\infer [VS-Catch] {(C,S,[]:E,\F,s) \Downarrow_{VS} (S',(h:t):E,F) \\ (C \wedge \mathit{catchCtx}(h:t),S'[(\GetCtx{C}{E},x) \rightarrow \mathit{combine}(h:t)],E,\F,s']) \Downarrow_{VS} (S'',E',F')} {(C,S,E,\F,\text{try}\ s\ \text{catch}\ x\ \text{in}\ s') \Downarrow_{VS} (S''[(\T,x) \rightarrow S'(\T,x)],E',F')}}
\def \BigVChcOne {\infer [VS-Chc1] {\mathit{sat}(\GetCtx{C}{E} \wedge A) \\ \mathit{sat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge A,S,E,\F,s) \Downarrow_{VS} (S',E',F) \\ (C \wedge \neg A,S',E',\F,s') \Downarrow_{VS} (S'',E'',F')} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S'',E'',F \wedge F')}}
\def \BigVChcTwo {\infer [VS-Chc2] {\mathit{unsat}(\GetCtx{C}{E} \wedge A) \\ \mathit{sat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge \neg A,S,E,\F,s') \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S',E',F)}}
\def \BigVChcThree {\infer [VS-Chc3] {\mathit{sat}(\GetCtx{C}{E} \wedge A) \\ \mathit{unsat}(\GetCtx{C}{E} \wedge \neg A) \\ (C \wedge A,S,E,\F,s) \Downarrow_{VS} (S',E',F)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S',E',F)}}
\def \BigVChcFour {\infer [VS-Chc4] {\mathit{unsat}(\GetCtx{C}{E} \wedge A) \\ \mathit{unsat}(\GetCtx{C}{E} \wedge \neg A)} {(C,S,E,\F,\tagtree{A}{s}{s'}) \Downarrow_{VS} (S,E,\T)}}

\begin{mathpar}
\BigVErr \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIfOne \and
\BigVIfTwo \and
\BigVIfThree \and
\BigVIfFour \and
\BigVWhileT \and
\BigVWhileF \and
\BigVThrow \and
\BigVTry \and
\BigVCatch
\end{mathpar}
\begin{mathpar}
\BigVChcOne \and
\BigVChcTwo \and
\BigVChcThree \and
\BigVChcFour
\end{mathpar}

>>>>

\begin{figure}
\begin{syntax}
\text{(Store)}\\
S &::=& S(x) & \textit{Lookup x in S} \\
\end{syntax}

\begin{mathpar}
\BigN \and
\BigVar \and
\BigAdd \and
\BigB \and
\BigNot \and
\BigAnd \and
\BigLess
\end{mathpar}
\caption{Big-step semantics of IMP expressions}
\label{fig:impexpr}
\end{figure}

@@

\begin{figure}
\begin{syntax}
\text{(Boolean literals)} \\
b &::=& \CCkeyw{true} \\
& | & \CCkeyw{false} \\
\text{(Conditions)}\\
C &::=& b & \textit{Boolean literal} \\
& | & D & \textit{Dimension name} \\
& | & \neg C & \textit{Negation} \\
& | & C \vee C & \textit{Disjunction} \\
& | & C \wedge C & \textit{Conjunction} \\
\text{(Arithmetic expressions)}\\
a &::=& n & \textit{Integer literal} \\
& | & x & \textit{Variable reference} \\
& | & a + a & \textit{Addition} \\
& | & \tagtree{C}{a}{a} & \textit{Choice} \\
\text{(Boolean expressions)} \\
e &::=& b & \textit{Boolean literal} \\
& | & \CCkeyw{not}\ e & \textit{Negation} \\
& | & e\ \CCkeyw{and}\ e & \textit{Conjunction} \\
& | & a < a & \textit{Less} \\
& | & \tagtree{C}{e}{e} & \textit{Choice} \\
\text{(Statements)} \\
s &::=& \CCkeyw{skip} & \textit{Noop} \\
& | & x \coloneqq a & \textit{Assignment} \\
& | & s\ ;\ s & \textit{Sequencing} \\
& | & \CCkeyw{if}\ e\ \CCkeyw{then}\ s\ \CCkeyw{else}\ s & \textit{Conditional} \\
& | & \CCkeyw{while}\ e\ \CCkeyw{do}\ s & \textit{Looping} \\
& | & \CCkeyw{throw}\ a & \textit{Throw error} \\
& | & \CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{in}\ s & \textit{Catch error} \\
& | & \tagtree{C}{s}{s} & \textit{Choice}
\end{syntax}
\caption{Syntax of Variational IMP with Effects}
\label{fig:vimpsyn}
\end{figure}

\def \BigVNum {\infer [VA-Num] { } {(C,S,n) \Downarrow_{VA} n}}
\def \BigVRef {\infer [VA-Ref] { } {(C,S,x) \Downarrow_{VA} S(C,x)}}
\def \BigVAdd {\infer [VA-Add] {(C,S,e) \Downarrow_{VA} u \\ (C,S,e') \Downarrow_{VA} u'} {(C,S,e+e') \Downarrow_{VA} \mathit{liftA2}(+,u,u')}}
\def \BigVAChcOne {\infer [VA-Chc1] {\mathit{sat}(C \wedge C') \\ \mathit{sat}(C \wedge \neg C') \\ (C \wedge C',S,a) \Downarrow_{VA} u \\ (C \wedge \neg C',S,a') \Downarrow_{VA} u'} {(C,S,\tagtree{C'}{a}{a'}) \Downarrow_{VA} \tagtree{C'}{u}{u'}}}
\def \BigVAChcTwo {\infer [VA-Chc2] {\mathit{unsat}(C \wedge C') \\ \mathit{sat}(C \wedge \neg C') \\ (C \wedge \neg C',S,a') \Downarrow_{VA} u} {(C,S,\tagtree{A}{a}{a'}) \Downarrow_{VA} u}}
\def \BigVAChcThree {\infer [VA-Chc3] {\mathit{sat}(C \wedge C') \\ \mathit{unsat}(C \wedge \neg C') \\ (C \wedge C',S,a) \Downarrow_{VA} u} {(C,S,\tagtree{C'}{a}{a'}) \Downarrow_{VA} u}}

\def \BigVB {\infer [VB-Bool] { } {(C,S,b) \Downarrow_{VB} b}}
\def \BigVNot {\infer [VB-Not] {(C,S,e) \Downarrow_{VB} v} {(C,S,\CCkeyw{not}\ e) \Downarrow_{VB} \mathit{liftA}(\neg,v)}}
\def \BigVAnd {\infer [VB-And] {(C,S,e) \Downarrow_{VB} v \\ (C,S,e') \Downarrow_{VB} v'} {(C,S,e\ \CCkeyw{and}\ e') \Downarrow_{VB} \mathit{liftA2}(\wedge,v,v')}}
\def \BigVLess {\infer [VB-Less] {(C,S,e) \Downarrow_{VA} v \\ (C,S,e') \Downarrow_{VA} v'} {(C,S,e<e') \Downarrow_{VB} \mathit{liftA2}(<,v,v')}}
\def \BigVBChcOne {\infer [VB-Chc1] {\mathit{sat}(C \wedge C') \\ \mathit{sat}(C \wedge \neg C') \\ (C \wedge C',S,e) \Downarrow_{VB} v \\ (C \wedge \neg C',S,e') \Downarrow_{VB} v'} {(C,S,\tagtree{C'}{e}{e'}) \Downarrow_{VB} \tagtree{C'}{v}{v'}}}
\def \BigVBChcTwo {\infer [VB-Chc2] {\mathit{unsat}(C \wedge C') \\ \mathit{sat}(C \wedge \neg C') \\ (C \wedge \neg C',S,e') \Downarrow_{VB} v} {(C,S,\tagtree{C'}{e}{e'}) \Downarrow_{VB} v}}
\def \BigVBChcThree {\infer [VB-Chc3] {\mathit{sat}(C \wedge C') \\ \mathit{unsat}(C \wedge \neg C') \\ (C \wedge C',S,e) \Downarrow_{VB} v} {(C,S,\tagtree{C'}{e}{e'}) \Downarrow_{VB} v}}

\begin{figure}
\begin{syntax}
\text{(Store)}\\
S &::=& S(C,x) & \textit{Lookup x in S with context C} \\
\text{(Variational integers)}\\
u &::=& n & \textit{Integer literal} \\
& | & \tagtree{C}{u}{u} & \textit{Choice} \\
\text{(Variational booleans)}\\
v &::=& b & \textit{Boolean literal} \\
& | & \tagtree{C}{v}{v} & \textit{Choice}
\end{syntax}

\begin{mathpar}
\BigVNum \and
\BigVRef \and
\BigVAdd \and
\BigVAChcOne \and
\BigVAChcTwo \and
\BigVAChcThree \and
\BigVB \and
\BigVNot \and
\BigVAnd \and
\BigVLess \and
\BigVBChcOne \and
\BigVBChcTwo \and
\BigVBChcThree
\end{mathpar}
\label{fig:vimpexpr}
\caption{Big-step semantics of Variational IMP with Effects expressions}
\end{figure}

\newcommand{\GetCtx}[2]{\mathit{ctx}(#1,#2)}
\def \gc {\GetCtx{C}{E}}
\def \T {\mathit{true}}
\def \F {\mathit{false}}
\def \BigVUnsat {\infer [V-Unsat] {\mathit{unsat}(\gc)} {(C,S,E,s) \Downarrow_{V} (S,E)}}
\def \BigVSat {\infer [V-Sat] {\mathit{sat}(\gc) \\ (C,S,E,s) \Downarrow_{VS} (S',E')} {(C,S,E,s) \Downarrow_{V} (S',E')}}
\def \BigVSkip {\infer[VS-Skip] { } {(C,S,E,\CCkeyw{skip}) \Downarrow_{VS} (S,E)}}
\def \BigVAssn {\infer[VS-Assn] {(\gc,S,a) \Downarrow_{VA} u} {(C,S,E,x \coloneqq a) \Downarrow_{VS} (S[(\gc,x) \rightarrow u],E)}}
\def \BigVSeq {\infer [VS-Seq] {(C,S,E,s) \Downarrow_{V} (S',E') \\ (C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,s\ ;\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVIf {\infer [VS-If] {(\gc,S,e) \Downarrow_{VB} v \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \mathit{whenFalse}(v),S',E') \Downarrow_{V} (S'',E'')} {(C,S,E,\CCkeyw{if}\ e\ \CCkeyw{then}\ s\ \CCkeyw{else}\ s') \Downarrow_{VS} (S'',E'')}}
\def \BigVWhileT {\infer [VS-WhileT] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{sat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E}) \\ (C \wedge \mathit{whenTrue}(v),S,E,s) \Downarrow_{VS} (S',E') \\ (C \wedge \mathit{whenTrue}(v),S',E',\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{VS} (S'',E'')} {(C,S,E,\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{VS} (S'',E'')}}
\def \BigVWhileF {\infer [VS-WhileF] {(\GetCtx{C}{E},S,e) \Downarrow_{VB} v \\ \mathit{unsat}(\GetCtx{C \wedge \mathit{whenTrue}(v)}{E})} {(C,S,E,\CCkeyw{while}\ e\ \CCkeyw{do}\ s) \Downarrow_{VS} (S,E)}}
\def \BigVThrow {\infer [VS-Throw] {(\GetCtx{C}{h:t},S,a) \Downarrow_{VA} u} {(C,S,h:t,\CCkeyw{throw}\ a) \Downarrow_{VS} (S,((C,u):h):t)}}
\def \BigVTry {\infer [VS-Try] {(C,S,[]:E,s) \Downarrow_{V} (S',[]:E)} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{in}\ s') \Downarrow_{VS} (S',E)}}
\def \BigVCatch {\infer [VS-Catch] {(C,S,[]:E,s) \Downarrow_{V} (S',h:E) \\ (\GetCtx{C}{(h:[])},S'[(\GetCtx{C}{E},x) \rightarrow \mathit{combine}(h)],E,s') \Downarrow_{V} (S'',E')} {(C,S,E,\CCkeyw{try}\ s\ \CCkeyw{catch}\ x\ \CCkeyw{in}\ s') \Downarrow_{VS} (S''[(\T,x) \rightarrow S'(\T,x)],E')}}
\def \BigVChc {\infer [VS-Chc] {(C \wedge C',S,E,s) \Downarrow_{V} (S',E') \\ (C \wedge \neg C',S',E',s') \Downarrow_{V} (S'',E'')} {(C,S,E,\tagtree{C'}{s}{s'}) \Downarrow_{VS} (S'',E'')}}

\begin{figure}
\begin{mathpar}
\BigVSat \and
\BigVUnsat \and
\BigVSkip \and
\BigVAssn \and
\BigVSeq \and
\BigVIf \and
\BigVWhileT \and
\BigVWhileF \and
\BigVThrow \and
\BigVTry \and
\BigVCatch \and
\BigVChc
\end{mathpar}
\caption{Big-step semantics of Variational Imp with Effects statements}
\label{fig:vimpstmt}
\end{figure}
